---
title: "Lecture 6: Point Level Models - Variograms & EDA"
output:
  revealjs::revealjs_presentation:
    theme: night
    center: true
    transition: none
    incremental: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
library(knitr)
library(ggplot2)
library(dplyr)
library(gstat)
library(sp)
library(datasets)
library(SemiPar)
#devtools::install_github("rstudio/leaflet")
library(leaflet)
```

# Class Intro

## Intro Questions 
- What is a variogram?
- How is a variogram useful?
- For Today:
    - How is EDA used for point referenced data?

# Variogram Creation


## Variogram Creation: How?

```{r, echo=F}
data(meuse)
meuse.small <- meuse %>% select(x, y, copper) %>% tbl_df()
meuse.small %>% head(15) %>% kable()
```

## Variogram Creation: Steps

1. Calculate distances between sampling locations
2. Choose grid for distance calculations
3. Calculate empirical semivariogram
$$\hat{\gamma}(d_k) = \frac{1}{2N(d_k)} \sum_{\boldsymbol{s}_i,\boldsymbol{s}_i, \in N(d_k)}\left[ \boldsymbol{Y}(\boldsymbol{s}_i) - \boldsymbol{Y}(\boldsymbol{s}_j)  \right]^2,$$
where $$N(d_k) = \{(\boldsymbol{s}_i,\boldsymbol{s}_j): ||\boldsymbol{s}_i -\boldsymbol{s}_j|| \in I_k\}$$ and $I_k$ is the $k^{th}$ interval.
4. Plot the semivariogram


## Variogram Creation: R function

```{r, eval = T}
data(meuse)
coordinates(meuse) = ~x+y
variogram(copper~1, meuse) %>% plot()
```

## Variogram Creation: How - Step 1

Calculate Distances between sampling locations

```{r, echo= t}
dist(meuse.small)
dist.mat <- dist(meuse.small %>% select(x,y))
```

## Variogram Creation: How - Step 2

Choose grid for distance calculations

```{r, echo= t}
cutoff <- max(dist.mat) / 3 # default maximum distance
num.bins <-  15
bin.width <- cutoff / 15
```

## Variogram Creation: How - Step 3

Calculate empirical semivariogram

```{r}
dist.sq <- dist(meuse.small$copper)^2

vario.dat <- data.frame(dist = as.numeric(dist.mat), diff = as.numeric(dist.sq)) %>% mutate(bin = floor(dist / bin.width) + 1) %>% filter(bin < 16) %>% group_by(bin) %>% summarize(emp.sv = .5 * mean(diff)) 
vario.dat

```

## Variogram Creation: How - Step 4

Plot empirical semivariogram
```{r}
vario.dat %>% ggplot(aes(x=bin, y= emp.sv)) + geom_point() + xlim(0,15) + ylim(0,800) + scale_x_continuous(breaks=c(5, 10, 15), labels =c('500','1000','1500')) + ggtitle('Empirical Semivariogram for Copper in Meuse Data Set') + xlab('distance (m)') + ylab('Empirical Semivariogram')
```

## Variogram Fitting

Now given this empirical semivariogram, how to we choose a semivariogram (and associated covariance structure) and estimate the parameters in that function??

```{r, fig.width= 6, fig.height = 4}
vario.dat %>% ggplot(aes(x=bin, y= emp.sv)) + geom_point() + xlim(0,15) + ylim(0,800) + scale_x_continuous(breaks=c(5, 10, 15), labels =c('500','1000','1500')) + ggtitle('Empirical Semivariogram for Copper in Meuse Data Set') + xlab('distance (m)') + ylab('Empirical Semivariogram')
```

## Variogram Fitting, cont..
- Empirical semivariograms can be computed and plotted, but they are subject to researcher choice: such as bin width and the number of points to display.
- The "data points" here is really a function of the observed random variable, but do not have an associated likelihood.
- Variogram fitting is very much an art and not a science.
- Eventually maximum likelihood or Bayesian methods will be used to estimate parameters in the covariance model directly.

# Exploratory Data Analysis

## EDA Overview
- Exploratory Data Analysis (EDA) is commonly used to explore and visualize data sets.
- EDA is not a formal analysis, but can inform modeling decisions.
- What are we interested in learning about with spatial data?

## Data Decomposition: Time Series
- In time series analysis, the first step in the EDA process was to decompose the observed data into a trend, seasonal cycles, and a random component.

```{r}
plot(decompose(AirPassengers))
```

## Data Decomposition: Spatial Data
- Similarly spatial data will be decomposed into the mean surface and the error surface.

- For example, elevation and distance from major bodies of water would be part of the mean surface for temperature.

- The mean surface is focused on the global, or first-order, behavior.

- The error surface captures local fluctuations, or second-order, behavior.

## Response Surface vs. Spatial Surface
- Spatial structure in the response surface and spatial structure in the error surface are not one-and-the-same.

- $E[(Y(\boldsymbol{s})-\mu)(Y(\boldsymbol{s'})-\mu)]$ vs. $E[(Y(\boldsymbol{s})-\mu(\boldsymbol{s}))(Y(\boldsymbol{s'})-\mu(\boldsymbol{s}))]$

- There are stationarity implications for considering the residual surface.

- Data sets contain two general types of useful information: spatial coordinates and covariates.

- Regression models will be used to build the mean surface.

## Spatial EDA Overview

1. Map of locations
2. Histrogram or other distributional figure
3. 3D scatterplot
4. General Regression EDA
5. Variograms and variogram clouds
6. Anistopic diagnostics

# Scallops Data Example

## 1. Map of Locations
```{r}
data(scallop)
scallop <- scallop %>% mutate(log.catch = log(tot.catch + 1))
leaflet(scallop) %>% addTiles() %>% addCircles()
```

## 1. Map of Locations - Takeaways

_Goal_: Understand the sampling approach

- Is this a grid?
- Are there directions that have larger distances?
- How large is the spatial extent?

## 2. Histogram
```{r}
scallop %>% ggplot(aes(x=tot.catch)) + geom_histogram(bins=70) + ggtitle('histogram of scallop catch') + xlab('Number of Scallops')
```

## 2. Histogram
```{r}
scallop %>% ggplot(aes(x=log.catch)) + geom_histogram(bins=70) + ggtitle('histogram of log scallop catch (+1)') + xlab('Number of Scallops')
```

## 2. Histogram - Takeaways

_Goal:_ Identify a sampling distribution for the data

- Continuous or discrete data
- A linear model approach will be used for the response
- Spatial structure can also be included in generalized linear models
- Outliers are worth investigating, but a data point that does not fit the assumed model should not automatically be eliminated

## 3. 3D scatterplot
```{r}
scallop %>%  ggplot(aes(x=longitude, y = latitude)) + geom_point(aes(color=tot.catch)) + scale_colour_gradient2()
```

## 3. 3D scatterplot
```{r}
scallop %>% ggplot(aes(x=longitude, y = latitude)) + geom_point(aes(color=log.catch)) + scale_colour_gradient2()
```

## 3. 3D scatterplot - Takeaways

_Goal:_ Examine the spatial pattern of the response

- Again, this is the response not the residual
- Can also think about a contour plot (using some interpolation method)

## 4. General Regression EDA

- Assessing relationship between variable of interest and covariate information
- No covariates are present in the scallops data


## 5. Variograms and variogram clouds: Exercise
Explore the code below: what are the differences in the three variograms?
```{r, echo = T, eval = F}
coordinates(scallop) = ~longitude+latitude
class(scallop)
scallop.sp <- scallop
proj4string(scallop.sp) <- CRS("+proj=longlat +datum=WGS84")  ## for example
scallop.utm <- spTransform(scallop.sp, CRS("+proj=utm +zone=18 ellps=WGS84"))

plot(variogram(log.catch~1, scallop))
plot(variogram(log.catch~1, scallop.sp))
plot(variogram(log.catch~1, scallop.utm))
```

## 5. Variograms 
```{r,  eval = T}
coordinates(scallop) = ~longitude+latitude
scallop.sp <- scallop
proj4string(scallop.sp) <- CRS("+proj=longlat +datum=WGS84")  ## for example
scallop.utm <- spTransform(scallop.sp, CRS("+proj=utm +zone=18 ellps=WGS84"))

plot(variogram(log.catch~1, scallop.utm))
```

## 5. Variogram Cloud
```{r}
plot(variogram(log.catch~1, scallop.utm, cloud=T))
```

## 5. Variograms and variogram clouds: Takeaways

_Goal:_ Visually diagnose spatial structure

## 6. Anisotropy

_Goal:_ Determine if direction influencs spatial structure

# Anisotropy

## Directional Variogram
- All of the variograms we have looked at are isotropic

```{r}
plot(variogram(log.catch~1, scallop.sp, alpha = c(0, 45, 90, 135)))
```

## Separable Correlations Functions

- If the differences in spatial structure are directly related to two coordinate sets, we can create a stationary, anistropic covariance function

- Let $$cor(Y(\boldsymbol{s + h}), Y(\boldsymbol{s})) = \rho_1(h_y) \rho_2(h_x),$$
where $\rho_1()$ and $\rho_2()$ are proper correlation functions.
- A scaling factor, $\sigma^2,$ can be used to create covariance.

## Geometric Anistropy

- Another solution is the class of geometric anisotropic covariance functions with 
$$C(\boldsymbol{s} - \boldsymbol{s'}) = \sigma^2 \rho((\boldsymbol{s} - \boldsymbol{s'})^TB(\boldsymbol{s} - \boldsymbol{s'})),$$
where $B$ is positive definite matrix and $\rho$ is a valid correlation function
- $B$ is often referred to as a transformation matrix which rotates and scales the coordinates, such that the resulting transformation can be simplified to a distance.

## Sill, Nugget, and Range Anisotropy
- Recall the sill is defined as $\lim_{d \rightarrow \infty} \gamma(d)$
- Let $\boldsymbol{h}$ be an arbitrary separation vector, that can be normalized as $\frac{\boldsymbol{h}}{||\boldsymbol{h}||}$
- If $\lim_{a \rightarrow \infty} \gamma(a \times \frac{\boldsymbol{h}}{||\boldsymbol{h}||})$ depends on $\boldsymbol{h}$, this is referred to as sill anisotropy.
- Similarly the nugget and range can depend on $\boldsymbol{h}$ and give nugget anisotropy and range anisotropy 

# Model Fitting

## Simulating Spatial Process

- Soon we will look at fitting models for spatial point data
- Simulating data gives a deeper understanding of the model fitting process
- Simulate a mean-zero, isotropic spatial process with a spherical covariance function


## Additional Resources
- [Meuse Data Tutorial](https://cran.r-project.org/web/packages/gstat/vignettes/gstat.pdf)
- [Textbook Data Sets](https://www.counterpointstat.com/hierarchical-modeling-and-analysis-for-spatial-data.html/)
